---
title: "Analyzing Trait Evolution with Approximate Bayesian Computation in TreEvo"
author: "David W. Bapst"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
plot(rnorm(100))
```

```{r,echo=FALSE}
library(paleotree)
data(RaiaCopesRule)
set.seed(2)

```

 Let's read in the trees from Raia et al 2015, AmNat
 following is taken from their supplemental appendix, available at AmNat

 they all appear to be trees dated to the last appearance times
 *and* specifically the end-boundary of the interval containing the last appearance


 ammonite genera


 now let's plot the tree

```{r}
plot(ladderize(ammoniteTreeRaia));axisPhylo()
```



this tree has polytomies - will have to deal with that

```{r}
tree<-multi2di(ammoniteTreeRaia) #randomly resolve tree
plot(ladderize(tree));axisPhylo()
```


plot with traits

 maybe use phytools rather than mine?


```{r}
plotTraitgram(tree=tree, trait=sutureComplexity,
	 conf.int=FALSE, main="Ammonite Suture Complexity")

plotTraitgram(tree=tree, trait=shellSize,
	 conf.int=FALSE, main="Ammonite Shell Diameter")
```

Let's use TreEvo!

```{r}

library(TreEvo)

# character data for doRun must be in matrix form
  # with rows labeled with taxon names

charData<-matrix(sutureComplexity,ncol=1)
rownames(charData)<-names(sutureComplexity)
```

Let's analyze it with ordinary BM

```{r}
results<-doRun_prc(
  phy = tree,
  traits = charData,
  intrinsicFn=brownianIntrinsic,
  extrinsicFn=nullExtrinsic,
  startingPriorsFns="normal",
  startingPriorsValues=matrix(c(mean(charData[,1]), sd(charData[,1]))),
  intrinsicPriorsFns=c("exponential"),
  intrinsicPriorsValues=matrix(c(10, 10), nrow=2, byrow=FALSE),
  extrinsicPriorsFns=c("fixed"),
  extrinsicPriorsValues=matrix(c(0, 0), nrow=2, byrow=FALSE),
  TreeYears=1000,
  standardDevFactor=0.2,
  plot=FALSE,
  StartSims=10,
  epsilonProportion=0.7,
  epsilonMultiplier=0.7,
  nStepsPRC=3,
  numParticles=20,
  jobName="examplerun_prc",
  stopRule=FALSE,
  multicore=FALSE,
  coreLimit=1
  )
```

Let's analyze it with ordinary BM with a bound

```{r}
resultsBound<-doRun_prc(
  phy = tree,
  traits = charData,
  intrinsicFn=boundaryMinIntrinsic,
  extrinsicFn=nullExtrinsic,
  startingPriorsFns="normal",
  startingPriorsValues=matrix(c(mean(charData[,1]), sd(charData[,1]))),
  intrinsicPriorsFns=c("exponential","normal"),
  intrinsicPriorsValues=matrix(c(10, 10, -10, 1), nrow=2, byrow=FALSE),
  extrinsicPriorsFns=c("fixed"),
  extrinsicPriorsValues=matrix(c(0, 0), nrow=2, byrow=FALSE),
  TreeYears=1000,
  standardDevFactor=0.2,
  plot=FALSE,
  StartSims=10,
  epsilonProportion=0.7,
  epsilonMultiplier=0.7,
  nStepsPRC=3,
  numParticles=20,
  jobName="examplerun_prc",
  stopRule=FALSE,
  multicore=FALSE,
  coreLimit=1
  )

```


















Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
