---
title: "Analyzing Trait Evolution with Approximate Bayesian Computation in TreEvo"
author: "David W. Bapst"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
#          fig_caption:: yes

vignette: >
  %\VignetteIndexEntry{Analyzing Trait Evolution with Approximate Bayesian Computation in TreEvo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r,echo=FALSE}
library(paleotree)
data(RaiaCopesRule)
set.seed(444)

```

 Let's read in the trees from Raia et al 2015, AmNat
 following is taken from their supplemental appendix, available at AmNat

 they all appear to be trees dated to the last appearance times
 *and* specifically the end-boundary of the interval containing the last appearance


 ammonite genera


 now let's plot the tree

```{r}
plot(ladderize(ammoniteTreeRaia));axisPhylo()
```



this tree has polytomies - will have to deal with that

```{r}
tree<-multi2di(ammoniteTreeRaia) #randomly resolve tree
tree<-addTermBranchLength(tree,0.01)

plot(ladderize(tree));axisPhylo()
```


plot with traits

 maybe use phytools rather than mine?


```{r}
plotTraitgram(tree=tree, trait=sutureComplexity,
	 conf.int=FALSE, main="Ammonite Suture Complexity")

plotTraitgram(tree=tree, trait=shellSize,
	 conf.int=FALSE, main="Ammonite Shell Diameter")
```

Let's use TreEvo!

```{r}

library(TreEvo)

# character data for doRun must be in matrix form
  # with rows labeled with taxon names

charData<-matrix(sutureComplexity,ncol=1)
rownames(charData)<-names(sutureComplexity)

```

Let's analyze it with ordinary BM

```{r}
results<-doRun_prc(
  phy = tree,
  traits = charData,
  intrinsicFn=brownianIntrinsic,
  extrinsicFn=nullExtrinsic,
  startingPriorsFns="normal",
  startingPriorsValues=matrix(c(mean(charData[,1]), sd(charData[,1]))),
  intrinsicPriorsFns=c("exponential"),
  intrinsicPriorsValues=matrix(c(10, 10), nrow=2, byrow=FALSE),
  extrinsicPriorsFns=c("fixed"),
  extrinsicPriorsValues=matrix(c(0, 0), nrow=2, byrow=FALSE),
  generation.times=10000,
  standardDevFactor=0.2,
  plot=FALSE,
  StartSims=10,
  epsilonProportion=0.7,
  epsilonMultiplier=0.7,
  nStepsPRC=3,
  numParticles=20,
  jobName="examplerun_prc",
  stopRule=FALSE,
  multicore=FALSE,
  coreLimit=1,
  verboseParticles=FALSE
  )
```

Let's analyze it with ordinary BM with a bound

```{r}
resultsBound<-doRun_prc(
  phy = tree,
  traits = charData,
  intrinsicFn=boundaryMinIntrinsic,
  extrinsicFn=nullExtrinsic,
  startingPriorsFns="normal",
  startingPriorsValues=matrix(c(mean(charData[,1]), sd(charData[,1]))),
  intrinsicPriorsFns=c("exponential","normal"),
  intrinsicPriorsValues=matrix(c(10, 10, -10, 1), nrow=2, byrow=FALSE),
  extrinsicPriorsFns=c("fixed"),
  extrinsicPriorsValues=matrix(c(0, 0), nrow=2, byrow=FALSE),
  generation.times=10000,
  standardDevFactor=0.2,
  plot=FALSE,
  StartSims=10,
  epsilonProportion=0.7,
  epsilonMultiplier=0.7,
  nStepsPRC=3,
  numParticles=20,
  jobName="examplerun_prc",
  stopRule=FALSE,
  multicore=FALSE,
  coreLimit=1,
  verboseParticles=FALSE
  )

```


```{r}

save.image(file="treevo_vignette_workspace.Rdata")

```












The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

