---
title: "Analyzing Trait Evolution with Approximate Bayesian Computation in TreEvo"
author: "David W. Bapst"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette

vignette: >
  %\VignetteIndexEntry{Analyzing Trait Evolution with Approximate Bayesian Computation in TreEvo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  
  
---

```{r,echo=FALSE}
# Control Box
set.seed(444)
multicore<-TRUE
coreLimit<-6
generation.time<-10000
```  

For this vignette, we will use the tree of ammonite genera, and the continuous character data for those ammonites, taken from Raia et al. (2015, American Naturalist). This data is taken from their supplemental appendix, available at the American Naturalist website, and is available as an example dataset in R package `paleotree`. 


```{r}
library(paleotree)
data(RaiaCopesRule)
```

```{r,echo=FALSE}
load("treevo_vignette_workspace_presaved.Rdata")
```

```{r}
ls()
```

The trees for the three groups examined in this paper all appear to be trees dated to the last appearance times (as opposed to the first appearance time) *and* specifically the end-boundary of the interval containing the last appearance.

First, let's plot the tree:

```{r}
plot(ladderize(ammoniteTreeRaia));axisPhylo()
```

This tree has polytomies - we will have to deal with that. Let's randomly resolve tree using `multi2di` from ape, and then use function `addTermBranchLength` from package `paleotree`

```{r}
tree<-multi2di(ammoniteTreeRaia)
tree<-addTermBranchLength(tree,0.01)
plot(ladderize(tree));axisPhylo()
```

Because we aren't adjust the internal edges at all, `multi2di` has just inserted a number of zero-length internal edges between internodes within the tree. This means that alternative resolutions of the tree (as multi2di is stochastic) won't create meaningfully different variance-covariance matrices, so we likely don't need to worry about resolving multiple trees in this case (because we're taking the trees used by Raia et al. at total face-value, which perhaps is the truly unsupported assumption).


Let's plot a traitgram of the data.

 maybe use phytools rather than mine?


```{r, echo=FALSE, fig.show='hold'}
plotTraitgram(tree=tree, trait=sutureComplexity,
	 conf.int=FALSE, main="Ammonite Suture Complexity")

plotTraitgram(tree=tree, trait=shellSize,
	 conf.int=FALSE, main="Ammonite Shell Diameter")
```

how small is the smallest branch length, even after extending terminal branches?

```{r}

brlen<-tree$edge.length
min(brlen)

# in years
min(brlen)*10^6

```


Let's use TreEvo!

```{r}

library(TreEvo)

# character data for doRun must be in matrix form
  # with rows labeled with taxon names

charData<-matrix(sutureComplexity,ncol=1)
rownames(charData)<-names(sutureComplexity)

```

Let's analyze it with ordinary BM

```{r eval=FALSE}


results<-doRun_prc(
  phy = tree,
  traits = charData,
  intrinsicFn=brownianIntrinsic,
  extrinsicFn=nullExtrinsic,
  startingPriorsFns="normal",
  startingPriorsValues=matrix(c(mean(charData[,1]), sd(charData[,1]))),
  intrinsicPriorsFns=c("exponential"),
  intrinsicPriorsValues=matrix(c(10, 10), nrow=2, byrow=FALSE),
  extrinsicPriorsFns=c("fixed"),
  extrinsicPriorsValues=matrix(c(0, 0), nrow=2, byrow=FALSE),
  generation.time=generation.time,
  standardDevFactor=0.2,
  plot=FALSE,
  StartSims=10,
  epsilonProportion=0.7,
  epsilonMultiplier=0.7,
  nStepsPRC=3,
  numParticles=20,
  jobName="typicalBMrun",
  stopRule=FALSE,
  multicore=multicore,
  coreLimit=coreLimit,
  verboseParticles=TRUE
  )

```

Let's analyze it with ordinary BM with a bound
  
```{r eval=FALSE}


resultsBound<-doRun_prc(
  phy = tree,
  traits = charData,
  intrinsicFn=boundaryMinIntrinsic,
  extrinsicFn=nullExtrinsic,
  startingPriorsFns="normal",
  startingPriorsValues=matrix(c(mean(charData[,1]), sd(charData[,1]))),
  intrinsicPriorsFns=c("exponential","normal"),
  intrinsicPriorsValues=matrix(c(10, 10, -10, 1), nrow=2, byrow=FALSE),
  extrinsicPriorsFns=c("fixed"),
  extrinsicPriorsValues=matrix(c(0, 0), nrow=2, byrow=FALSE),
  generation.time=generation.time,
  standardDevFactor=0.2,
  plot=FALSE,
  StartSims=10,
  epsilonProportion=0.7,
  epsilonMultiplier=0.7,
  nStepsPRC=3,
  numParticles=20,
  jobName="BMwithBoundRun",
  stopRule=FALSE,
  multicore=multicore,
  coreLimit=coreLimit,
  verboseParticles=FALSE
  )


```


```{r, eval=FALSE, echo=FALSE}

save.image(file="treevo_vignette_workspace.Rdata")

```

